\MakeUppercase{Linear versus centred Colouring Numbers}\thanks{This research was partly funded by NSERC.}

### 2022-04-08 lin-vs-cen.tex
Pffft

\begin{abstract}
  We prove that the linear colouring number of any $k\times k$ pseudogrid is $\Omega(k)$.  By an argument of Kun et al, this result gives a tighter upper bound on the treedepth of a graph as a function of its linear colouring number and gives further evidence in support of their conjecture that the treedepth of any graph is upper bounded by a linear function of its linear colouring number.
\end{abstract}

@@

The question of upper bounding $\chicen(G)$ by some function of $\chilin(G)$ was considered by \citet{kun.obrien.ea:polynomial,czerwinski.nadara.ea:improved} who prove:

@@

Their proof of \cref{kun_obrien_general} has three steps:
\begin{enumerate}
  \item A theorem of \citet{czerwinski.nadara.ea:improved} shows that, if $\chicen(G)\ge k^{19}\log^{q} k$ then $G$ contains a subcubic tree of treedepth $\Omega(k)$, or $\tw(G)\in\Omega(k^{18}\log^q k)$.\footnote{The result of \citet{czerwinski.nadara.ea:improved} is, of course, more general:  If $\chicen(G)\ge a\times b$, then $G$ contains an $\Omega(a)\times\Omega(a)$ grid minor or $G$ contains a subcubic tree of treedepth $\Omega(b)$. This is just an application of their result with $a=k^{18}\log^q k$ and $b=k$.}  In the former case, an asymptotically optimal result of \citet{kun.obrien.ea:polynomial} on subcubic trees completes the proof, so we are left with the case where $\tw(G)\in\Omega(k^{18}\log^q k)$.
  \item The current-best version of the Excluded Grid Theorem due to \citet{chuzhoy.tan:towards} shows that, if $\tw(G)\in\Omega(k^{18}\log^q k)$ (for a particular fixed positive $q$), then $G$ contains an $\Omega(k^2)\times \Omega(k^2)$ grid minor.
  \item A technical lemma \cite[Lemma~5]{kun.obrien.ea:polynomial} shows that, if $G$ contains a $\Omega(k^2)\times \Omega(k^2)$ grid minor, then $\chilin(G)\in\Omega(k)$.
\end{enumerate}

In the current work, we spend considerable effort to improve the technical lemma in Point~3 to show that a $\Omega(k)\times\Omega(k)$ grid minor is sufficient to guarantee that $\chilin(G)\in\Omega(k)$. This improves the exponent in \cref{kun_obrien_general} from $19$ to $10$, yielding the following improvement to \cref{kun_obrien_general}:
\begin{thm}\label{kun_obrien_general2}
  For any graph $G$, $\chicen(G)\in O(\chilin^{10}(G))$.
\end{thm}

\todo[inline]{Open problem here: Is there anything special about the relationship between treedepth and grid minors in planar graphs or product structured graphs?  It's probably worth reading \citet{czerwinski.nadara.ea:improved} to see how their proof works.}

@@

In this paper, all graphs are simple and undirected. For a graph $G$, $V(G)$ denotes the vertex set of $G$, $E(G)$ denotes the edge set of $G$ and $\VE(G):=V(G)\cup E(G)$ denotes the set of vertices and edges of $G$. For a vertex $v\in V(G)$,  $N_G(v):=\{w\in V(G):vw\in E(G)\}$  denotes the open neighbourhood of $v$ in $G$ and for any set $S\subseteq V(G)$, $N_G(S):=\{w\in V(G):vw\in E(G),\, v\in S,\, w\not\in S\}$.

For a  bipartite graph $H$, the two parts of $V(H)$ are denoted by $L(H)$ and $R(H)$ and we use the convention of writing an edge $xy$ so that its first endpoint $x$ is in $L(H)$ and its second endpoint $y$ is in $R(H)$.  A \defin{matching} $M$ in a bipartite graph $H$ is a subgraph of $H$ in which each vertex has degree at most $M$.  We say that $M$ \defin{saturates} a set $S\subseteq V(H)$ if $\deg_M(v)=1$ for each $v\in S$.  We make use of (the difficult half of) Hall's Marriage Theorem (see, for example \citet[Theorem~2.1.2]{diestel:graph}):

@@

\cref{d_hall} can be deduced by applying adding $d-1$ twins $x_2,\ldots,x_d$ for each vertex $x\in L(H)$, i.e., by applying \cref{hall} on the graph $H':=H\cup\{x_iy:x\in L(H),\, i\in\{2,\ldots,d\},\, y\in N_H(x)\}$.

@@

For positive integers $a$ and $b$, the \defin{$a\times b$ grid} $G_{a\times b}$ is the graph with vertex set $V(G_{a\times b}):=\{1,\ldots,a\}\times\{1,\ldots,b\}$ and that contain an edge with endpoints $(i_1,j_1)$ and $(i_2,j_2)$ if and only if $|i_1-i_2|+|j_1-j_2|=1$.  Such an edge is \defin{vertical} if $i_1=i_2$ and \defin{horizontal} if $j_1=j_2$.  For each $i\in\{1,\ldots,a\}$, the \defin{$i$th column} of $G_{a\times b}$ is the vertex set $\{(i,1),\ldots,(i,b)\}$ and, for each $j\in\{1,\ldots,b\}$, the $j$th row is the vertex set $\{(1,j),\ldots,(a,j)\}$.  For any integer $0\le r\le\min\{a,b\}/2$, the \defin{r-interior} of $G_{a\times b}$ defined as $\interior_r(G_{a\times b}):=G_{a\times b}[\{1+r,\ldots,a-r\}\times\{1+r,\ldots,b-r\}]$.  For $r\ge\min\{a,b\}/2$, $\interior_r(G_{a\times b})$ is the empty graph.

@@

In this work, we need to work with pseudogrids.  Refer to \cref{pseudogrid_fig}.  An \defin{$a\times b$-pseudogrid} is any graph that can be obtained from $G_{a\times b}$ in the following way:
\begin{compactitem}
  \item Replace each edge $vw$ of $G_{a\times b}$ with a path $\overline{P}_{vw}$ whose endpoints are $v$ and $w$.  (In other words, $\overline{P}_{vw}$ is a path obtained by subdividing the edge $vw$ zero or more times.)
  \item Replace each degree-$4$ vertex $(i,j)$ of $G_{a\times b}$ with a path $P_v$. If $P_v$ has only one vertex then the unique vertex of $P_v$ is adjacent to each of $(i\pm 1,j\pm 1)$ and this has no effect on the underlying graph.  Otherwise, $P_v$ has two endpoints $p$ and $q$, each of which is adjacent to two vertices among $(i\pm 1,j\pm 1)$.  It is useful to consider three possible cases:
  \begin{compactenum}[(Q1)]
    \item \label{q_i} $p$ is adjacent to $\{(i-1,j), (i,j-1)\}$ and $q$ is adjacent to $(i+1,j),(i,j+1)$;
    \item \label{q_ii} $p$ is adjacent to $\{(i,j+1), (i-1,j)\}$ and $q$ is adjacent to $(i,j-1),(i+1,j)$; or
    \item \label{q_iii} $p$ is adjacent to $\{(i-1,j),(i+1,j)\}$ and $q$ is adjacent to $\{(i,j-1),(i,j+1)\}$.
  \end{compactenum}
\end{compactitem}

@@

Each row $R':=v_1,\ldots,v_a$ of $G_{a\times b}$ corresponds naturally to a path $R$ of $G$. The path $R$ contains $V(\overline{P}_{v_iv_{i+1}})$ for each $i\in\{1,\ldots,a-1\}$.  However, for $i\in\{2,\ldots,a-1\}$ $R$ may or may not contain $V(P_{v_i})$.  In particular, if $P_{v_i}$ was created using (Q\ref{q_iii}) then $R$ does not contain any internal vertices in $P_{v_i}$. Similarly, a column $C':=v_1,\ldots,v_b$ of $G_{a\times b}$ corresponds to a path that contains $V(\overline{P}_{v_jv_{j+1}})$ for each $j\in\{1,\ldots,b-1\}$. This correspondence allows us to talk about the rows and columns of $G$, which we will do immediately.

As part of our proof, we use the operation of \defin{deleting} a row (or column) $R$ of $G$, which works as follows:  We remove all edges of $R$ from $G$ and eliminate any isolated vertices.  If this produces vertices of degree $1$ (which happens when $R$ is the first or last row of $G$ or when $R=v_1,\ldots,v_a$ does not contains $P_{v_i}$ for some $i\in\{1,\ldots,r\}$) then we repeatedly remove degree-$1$ vertices until none remain.  If $G$ is an $a\times b$ pseudogrid and we delete some row $R$, then the resulting graph is an $a\times (b-1)$ pseudogrid.  Similarly, if we delete column $C$ of $G$, then the resulting graph is a $(a-1)\times b$ pseudogrid.

@@

The most challenging aspect of this proof is the construction of $S_2$, which requires the use of the Local Lemma (\cref{weighted_lovasz}) to ensure that no vertex chosen to take part in $S_2$ is close to any other vertex in $S_2$.  The difficulty is illustrated in \cref{outline} by the cluster of three points of $S_1$ in the top right corner whose boxes overlap.  These three points of $S_1$ are well-separated, but choosing one point from each of the three boxes to take part in $S_2$ could result in three points of $S_2$ being very close to each other.  The local lemma is used to show that, in situations like this, we can still obtain two vertices of $S_2$ from each missing colour class while avoiding taking more than one vertex of $S_2$ from these three boxes.

@@

\begin{proof}
  The proof is by induction on $|\varphi(V(G))|$, the number of colours used by the colouring $\varphi$.  If there exists no $A\subseteq \varphi(V(G))$ with $|\varphi_{\mathcal{P}}^{-1}(A)\cap\interior_r(G_k)| < d|A|$ then taking $G':=G$ and $\mathcal{P}':=\mathcal{P}$ satisfies the requirements of the lemma.  Otherwise, we will remove a set $R$ of rows and a set $C$ of columns from $G$ with $|R|=|C|\le d|A|+2r \le (d+2r)|A|$ to eliminate all vertices with colours in $A$, as follows:
  \begin{compactitem}
    \item For each $v:=(i,j)\in V(\interior_r(G_{k}))$ with $A\cap\varphi_\mathcal{P}(v)\neq\emptyset$, we include row $j$ in $R$ and column $i$ in $C$.
    \item For each horizontal edge $vw\in E(\interior_r(G_{k}))$ with $A\cap\varphi_\mathcal{P}(vw)\neq\emptyset$, we include the row of $G$ that contains $P_{vw}$ in $R$.
    \item For each vertical edge $vw\in E(\interior_r(G_{k}))$ with $A\cap\varphi_\mathcal{P}(vw)\neq\emptyset$, we include the row of $G$ that contains $P_{vw}$ in $R$.
    \item We add the first and last $r$ rows to $R$ and the first and last $r$ columns to $C$.
    \item Finally, we add arbitrary rows to $R$ or columns to $C$ to ensure that $|R|=|C|$.
  \end{compactitem}
  At this point $|R|=|C|\le (d+2r)|A|$ and we remove all rows in $R$ and all columns in $C$ from $G$ to obtain a $k_0\times k_0$ pseudogrid $G_0$ with $k_0\ge k-d|A|$ and such that $\varphi(V(G_0))\cap A=\emptyset$.  In particular, $|\varphi(V(G_0))|\le |\varphi(V(G))|-|A|$.

@@

\begin{lem}\label{one_colour_per_object}
  Let $d,r>1$ be integers, let $G$ be a $k\times k$ pseudogrid with grid-partition $\mathcal{P}$, and let $\varphi:V(G)\to\{1,\ldots,c\}$ be a vertex colouring of $G$ such that, for any $A\subseteq\varphi(V(G))$, $|\varphi_{\mathcal{P}'}^{-1}(A)\cap\interior_r(G_k)| \ge d\cdot|A|$. Then there exists a colouring $\phi:\VE(G_{k})\to\{1,\ldots,c,\perp\}$ with the following properties:
  \begin{compactenum}[(i)]
    \item For each $\mu\not\in\interior_r(G_{k})$, $\phi(\mu)=\perp$.
    \item For each $\mu\in\VE(G_{k})$, $\phi(\mu)=\perp$ or $\phi(\mu)\in\varphi_\mathcal{P}(\mu)$.
    \item For each $\alpha\in\varphi(V(G))$,
    $|\phi^{-1}(\alpha)|\ge d$.
    % $|\{\mu\in\VE(G_{k}):\phi(\mu)=\alpha\}|\ge d$.
  \end{compactenum}
\end{lem}

@@

Next we show that, given a sufficiently `well-separated' set $S$ of pairs of vertices in $G$, we can always find a path in $G$ that contains every vertex in $S$. For this, we need some definitions of boxes in $G$ and in $G_k$.

@@

We extend these definitions to vertices of a $k\times k$ pseudogrid $G$ with grid-partition $\mathcal{P}:=\{V(P_\mu):\mu\in\VE(G_k)\}$ as follows. For any $\mu\in V(G_k)$, define
\[
   \tilde{B}_r(\mu) := \bigcup_{\nu\in \VE(G_r(\mu))} V(P_\nu)
\]
and, for any $v\in V(G)$, let $\tilde{B}_r(v):=\tilde{B}_r(\mu_v)$
where $\mu_v$ denotes the unique element in $\VE(G_k)$ such that $v\in P_{\mu_v}$.\footnote{Technically, the notation for $\tilde{B}_r(v)$ should include the partition $\mathcal{P}$, but we omit this since there will never be any ambiguity as to which partition is being used.}
% For any $v\in V(G)$, define $\tilde{G}_r(v):=G[\tilde{B}_r(v)]$.

\begin{lem}\label{pick_up_two}
  Let $G$ be a $a\times a$ pseudogrid with $a\ge??$,  let $s,v,w,t$ be vertices of $G$ with $s$ in the first column of $G$, neither $v$ nor $w$ in the first or last column of $G$ and $t$ in the last column of $G$.  Then $G$ contains a path $P$ with endpoints $s$ and $t$ that contains $v$ and $w$.
\end{lem}

\begin{proof}
  Let $\mathcal{P}:\{P_\mu:\mu\in \VE(G_k)\}$ be a grid partition of $G$ and let $\mu_v,\mu_w\in \VE(G_k)$ be such that $v\in V(P_{\mu_v})$ and $w\in V(P_{\mu_w})$.
  \todo[inline]{Finish this.}
\end{proof}

@@

  Let $X:=X_1\cup X_2\cup X_3$.  By construction, $X$ satisfies (\ref{covers_s}), so it only remains to show that $X$ satisfies (\ref{two_per_box}) and (\ref{disjoint_boxes}).  That $X$ satisfies (\ref{two_per_box}) follows easily from (\ref{covers_s}) and the fact that each $x\in X$ covers at most two elements of $S$.

  To see that $X$ satisifies (\ref{disjoint_boxes}), observe that if $x\in X$ covers $v\in S$ then $B_{p+1}(x)\subseteq B_{2p+2}(v)$. Therefore, if $x\in X$ covers $v\in S$ and $y\in X$ covers $w\in S$ and $B_{p+1}(x)\cap B_{p+1}(y)\neq\emptyset$ then $w\in B_{4p+4}(v)$ or $v\in B_{4p+4}(w)$.  Since $r-1 \ge 4p+4$, this implies that $v=w$ or $v$ and $w$ are an $(r-1)$-pair.  In the later case, $x=y$ and there is nothing to prove. Otherwise $v,w\in S_2$ or $v,w\in S_3$.  If $v,w\in S_2$ then, again $x=y$ and there is nothing to prove.  If $v,w\in S_3$ then $x$ and $y$ are specifically chosen so that $B_{p+1}(x)$ and $B_{p+1}(y)$ are disjoint.
\end{proof}

@@

\begin{proof}
  Apply \cref{make_disjoint} with $p=\floor{(r-5)/4}$ to obtain the set $X$.  Now, consider a `snake-like' path $P_0$ that contains every row $i$ with $i\equiv 1\pmod 2p+1$ and, aside from those rows contains only vertices of columns $1$ and $k$.   For each $x\in X$, this path intersects exactly one row of $B_{p+1}(x)$.  For each $x\in X$ we use \cref{pick_up_two} to replace $V(P_0)\cap B_{p+1}(x)$ with a path that is contained in $B_{p+1}(x)$ and contains the (at most two) vertices of $B_{p+1}(x)\cap S$.  After doing this for each $x\in X$ we obtain a path $P$ that contains every vertex in $S$. 
\end{proof}

@@

\begin{proof}
  For simplicity, assume that $G_k$ is embedded in $G_{k+2r}$ with the elements of $Q$ all contained in $\VE(\interior_{2r}(G_{k+2r})$.  In particular, for each $\mu_1\in Q$, this implies that $|B_{\rho}(\nu)| =\vol{\rho}$ for any $\rho\in\{0,\ldots,2r\}$.

@@

\begin{lem}\label{doubled_colour_set}
  For every integer $r\ge 1$ there exists an integer $d\in O(r^4)$ such that the following is true, for every integer $k\ge 1$.
  Let $G$ be a $k\times k$ pseudogrid with grid-partition $\mathcal{P}:=\{V(P_\mu):\mu\in\VE(G_k)\}$ and let $\varphi$ be a vertex colouring of $G$ with the property that, for each $A\subseteq\varphi(V(G))$, $|\varphi_{\mathcal{P}}^{-1}(A)\cap\interior_r(G)|\ge d|A|$.
  Then there exists $S\subseteq V(G)$ such that
  \begin{compactenum}[(i)]
    \item \label{hits_both} for each $\alpha\in\varphi(V(G))$, $|\varphi^{-1}(\alpha)\cap S|= 2$; and
    \item \label{spread_out} for each $v\in S$, $|\tilde{B}_r(v)\cap S|\le 2$ for each $v\in S$.
  \end{compactenum}
\end{lem}

>>>>

\begin{abstract}
  We prove that the linear colouring number of any $k\times k$ pseudogrid is $\Omega(k)$.  By an argument of Kun et al (Algorithmica, 2021), this result gives a tighter upper bound on the treedepth of a graph as a function of its linear colouring number and gives further evidence in support of their conjecture that the treedepth of any graph is upper bounded by a linear function of its linear colouring number.
\end{abstract}

@@

Motivated by finding efficiently-computable approximations of treedepth in bounded expansion classes, linear colourings were introduced by \citet{kun.obrien.ea:polynomial} who were able to establish a polynomial upper bound on $\chicen(G)$ as a function of $\chilin(G)$.  Combined with recent improvements by \citet{czerwinski.nadara.ea:improved} on the relationship between treedepth and treewidth, they obtain the following result:

@@

Their proof of \cref{kun_obrien_general} has three main ingredients:
\begin{enumerate}
  \item A theorem of \citet{czerwinski.nadara.ea:improved} shows that, if $\chicen(G)\ge k^{19}\log^{q} k$ then $G$ contains a subcubic tree of treedepth $\Omega(k)$, or $\tw(G)\in\Omega(k^{18}\log^q k)$.\footnote{The result of \citet{czerwinski.nadara.ea:improved} is, of course, more general:  If $\chicen(G)\ge a\times b$, then $G$ contains an $\Omega(a)\times\Omega(a)$ grid minor or $G$ contains a subcubic tree of treedepth $\Omega(b)$. This is just an application of their result with $a=k^{18}\log^q k$ and $b=k$.}  In the former case, an asymptotically optimal result of \citet{kun.obrien.ea:polynomial} on subcubic trees completes the proof, so we are left with the case where $\tw(G)\in\Omega(k^{18}\log^q k)$.
  \item The current-best version of the Excluded Grid Theorem due to \citet{chuzhoy.tan:towards} shows that, if $\tw(G)\in\Omega(k^{18}\log^q k)$ (for a particular fixed positive $q$), then $G$ contains an $\Omega(k^2)\times \Omega(k^2)$ grid minor.
  \item A technical lemma \cite[Lemma~5]{kun.obrien.ea:polynomial} shows that, if $G$ contains a $\Omega(k^2)\times \Omega(k^2)$ grid minor, then $\chilin(G)\in\Omega(k)$.
\end{enumerate}

In the current work, we spend considerable effort to improve the technical lemma in Point~3 to show that a $\Omega(k)\times\Omega(k)$ grid minor is sufficient to guarantee that $\chilin(G)\in\Omega(k)$. This improves the exponent in \cref{kun_obrien_general} from $19$ to $10$, yielding the following improvement to \cref{kun_obrien_general}:
\begin{thm}\label{kun_obrien_general2}
  For any graph $G$, $\chicen(G)\in (\chilin(G))^{10}\log(\chilin(G))^{O(1)}$.
\end{thm}

% \todo[inline]{Open problem here: Is there anything special about the relationship between treedepth and grid minors in planar graphs or product structured graphs?  It's probably worth reading \citet{czerwinski.nadara.ea:improved} to see how their proof works.}

@@

In this paper, all graphs are simple and undirected. For a graph $G$, $V(G)$ denotes the vertex set of $G$, $E(G)$ denotes the edge set of $G$ and $\VE(G):=V(G)\cup E(G)$ denotes the set of vertices and edges of $G$. For a vertex $v\in V(G)$,  $N_G(v):=\{w\in V(G):vw\in E(G)\}$  denotes the open neighbourhood of $v$ in $G$ and for any set $S\subseteq V(G)$, $N_G(S):=\{w\in V(G):vw\in E(G),\, v\in S,\, w\not\in S\}$.  We use $\deg_G(v):=|N_G(v)|$ to denote the degree of the vertex $v$ in the graph $G$.

For a  bipartite graph $H$, the two parts of $V(H)$ are denoted by $L(H)$ and $R(H)$ and we use the convention of writing an edge $xy$ so that its first endpoint $x$ is in $L(H)$ and its second endpoint $y$ is in $R(H)$.  A \defin{matching} $M$ in a bipartite graph $H$ is a subgraph of $H$ in which each vertex has degree at most $1$.  We say that $M$ \defin{saturates} a set $S\subseteq V(H)$ if $\deg_M(v)=1$ for each $v\in S$.  We make use of (the difficult half of) Hall's Marriage Theorem (see, for example \citet[Theorem~2.1.2]{diestel:graph}):

@@

\cref{d_hall} can be deduced by adding $d-1$ twins $x_2,\ldots,x_d$ for each vertex $x\in L(H)$, i.e., by applying \cref{hall} on the graph $H':=H\cup\{x_iy:x\in L(H),\, i\in\{2,\ldots,d\},\, y\in N_H(x)\}$.

@@

For positive integers $a$ and $b$, the \defin{$a\times b$ grid} $G_{a\times b}$ is the graph with vertex set $V(G_{a\times b}):=\{1,\ldots,a\}\times\{1,\ldots,b\}$ and that contain an edge with endpoints $(i_1,j_1)$ and $(i_2,j_2)$ if and only if $|i_1-i_2|+|j_1-j_2|=1$.  Such an edge is \defin{vertical} if $i_1=i_2$ and \defin{horizontal} if $j_1=j_2$.  For each $i\in\{1,\ldots,a\}$, the \defin{$i$th column} of $G_{a\times b}$ is the vertex set $\{(i,1),\ldots,(i,b)\}$ and, for each $j\in\{1,\ldots,b\}$, the \defin{$j$th row} is the vertex set $\{(1,j),\ldots,(a,j)\}$.  For any integer $0\le r\le\min\{a,b\}/2$, the \defin{r-interior} of $G_{a\times b}$ defined as $\interior_r(G_{a\times b}):=G_{a\times b}[\{1+r,\ldots,a-r\}\times\{1+r,\ldots,b-r\}]$.  For $r\ge\min\{a,b\}/2$, $\interior_r(G_{a\times b})$ is the empty graph.

@@

Refer to \cref{pseudogrid_fig}.  An \defin{$a\times b$-pseudogrid} is any graph that can be obtained from $G_{a\times b}$ in the following way:
\begin{compactitem}
  \item Replace each edge $vw$ of $G_{a\times b}$ with a path $\overline{P}_{vw}$ whose endpoints are $v$ and $w$.  (In other words, $\overline{P}_{vw}$ is a path obtained by subdividing the edge $vw$ zero or more times.)
  \item Replace each degree-$4$ vertex $(i,j)$ of $G_{a\times b}$ with a non-empty path $P_v$. If $P_v$ has only one vertex then the unique vertex of $P_v$ is adjacent to each of $(i\pm 1,j\pm 1)$ and this has no effect on the underlying graph.  Otherwise, $P_v$ has two endpoints $p$ and $q$, each of which is adjacent to two vertices among $(i\pm 1,j\pm 1)$.  It is useful to consider three possible cases:
  \begin{compactenum}[(Q1)]
    \item \label{q_i} $p$ is adjacent to $\{(i-1,j), (i,j-1)\}$ and $q$ is adjacent to $(i+1,j),(i,j+1)$;
    \item \label{q_ii} $p$ is adjacent to $\{(i,j+1), (i-1,j)\}$ and $q$ is adjacent to $(i,j-1),(i+1,j)$; or
    \item \label{q_iii} $p$ is adjacent to $\{(i-1,j),(i+1,j)\}$ and $q$ is adjacent to $\{(i,j-1),(i,j+1)\}$.
  \end{compactenum}
\end{compactitem}

@@

Each row $R':=v_1,\ldots,v_a$ of $G_{a\times b}$ corresponds naturally to a path $R$ of $G$. The path $R$ contains $V(\overline{P}_{v_iv_{i+1}})$ for each $i\in\{1,\ldots,a-1\}$.  However, for $i\in\{2,\ldots,a-1\}$ $R$ may or may not contain $V(P_{v_i})$.  In particular, if $P_{v_i}$ was created using (Q\ref{q_iii}) then $R$ does not contain any internal vertices in $P_{v_i}$. Similarly, a column $C':=v_1,\ldots,v_b$ of $G_{a\times b}$ corresponds to a path $C$  in $G$ that contains $V(\overline{P}_{v_jv_{j+1}})$ for each $j\in\{1,\ldots,b-1\}$. This correspondence allows us to talk about the rows and columns of $G$, which we will do immediately.

As part of our proof, we use the operation of \defin{deleting} a row (or column) $R$ of $G$, which works as follows:  We remove all edges of $R$ from $G$ and eliminate any isolated vertices.  If this produces vertices of degree $1$ (which happens when $R$ is the first or last row of $G$ or when $R=v_1,\ldots,v_r$ does not contains $P_{v_i}$ for some $i\in\{1,\ldots,r\}$) then we repeatedly remove degree-$1$ vertices until none remain.  If $G$ is an $a\times b$ pseudogrid and we delete some row $R$, then the resulting graph is an $a\times (b-1)$ pseudogrid.  Similarly, if we delete column $C$ of $G$, then the resulting graph is a $(a-1)\times b$ pseudogrid.

@@

The most challenging aspect of this proof is the construction of $S_2$, which requires the use of the Local Lemma (\cref{weighted_lovasz}) to ensure that no vertex chosen to take part in $S_2$ is close to any other vertex in $S_2$.  The difficulty is illustrated in \cref{outline} by the cluster of three points of $S_1$ in the top right corner whose boxes overlap.  These three points of $S_1$ are well-separated, but choosing one point from each of the three boxes to take part in $S_2$ could result in three points of $S_2$ being very close to each other.  The local lemma is used to show that, in situations like this, we can still obtain two vertices from each missing colour class to put into $S_2$ without having to take more than one vertex from these three boxes.

@@

\begin{proof}
  The proof is by induction on $|\varphi(V(G))|$, the number of colours used by the colouring $\varphi$.  If there exists no $A\subseteq \varphi(V(G))$ with $|\varphi_{\mathcal{P}}^{-1}(A)\cap\interior_r(G_k)| < d|A|$ then taking $G':=G$ and $\mathcal{P}':=\mathcal{P}$ satisfies the requirements of the lemma.  Otherwise, we will remove a set $R$ of rows and a set $C$ of columns from $G$ with $|R|=|C|\le d|A|+2r \le (d+2r)|A|$ to eliminate all vertices with colours in $A$, as follows:
  \begin{compactitem}
    \item For each vertex $v:=(i,j)\in V(\interior_r(G_{k}))$ with $A\cap\varphi_\mathcal{P}(v)\neq\emptyset$, we include row $j$ in $R$ and column $i$ in $C$.
    \item For each horizontal edge $vw\in E(\interior_r(G_{k}))$ with $A\cap\varphi_\mathcal{P}(vw)\neq\emptyset$, we include the row of $G$ that contains $P_{vw}$ in $R$.
    \item For each vertical edge $vw\in E(\interior_r(G_{k}))$ with $A\cap\varphi_\mathcal{P}(vw)\neq\emptyset$, we include the row of $G$ that contains $P_{vw}$ in $R$.
    \item We add the first and last $r$ rows to $R$ and the first and last $r$ columns to $C$.
    \item Finally, we add arbitrary rows to $R$ or columns to $C$ to ensure that $|R|=|C|$.
  \end{compactitem}
  At this point $|R|=|C|\le (d+2r)|A|$ and we remove all rows in $R$ and all columns in $C$ from $G$ to obtain a $k_0\times k_0$ pseudogrid $G_0$ with $k_0\ge k-(d+2r)|A|$ and such that $\varphi(V(G_0))\cap A=\emptyset$.  In particular, $|\varphi(V(G_0))|\le |\varphi(V(G))|-|A|$.

@@

\begin{lem}\label{one_colour_per_object}
  Let $d,r>1$ be integers, let $G$ be a $k\times k$ pseudogrid with grid-partition $\mathcal{P}$, and let $\varphi:V(G)\to\{1,\ldots,c\}$ be a vertex colouring of $G$ such that, for any $A\subseteq\varphi(V(G))$, $|\varphi_{\mathcal{P}}^{-1}(A)\cap\interior_r(G_k)| \ge d\cdot|A|$. Then there exists a colouring $\phi:\VE(G_{k})\to\{1,\ldots,c,\perp\}$ with the following properties:
  \begin{compactenum}[(i)]
    \item For each $\mu\not\in\interior_r(G_{k})$, $\phi(\mu)=\perp$.
    \item For each $\mu\in\VE(G_{k})$, $\phi(\mu)=\perp$ or $\phi(\mu)\in\varphi_\mathcal{P}(\mu)$.
    \item For each $\alpha\in\varphi(V(G))$,
    $|\phi^{-1}(\alpha)|\ge d$.
    % $|\{\mu\in\VE(G_{k}):\phi(\mu)=\alpha\}|\ge d$.
  \end{compactenum}
\end{lem}

@@

Next we show that, given a sufficiently `well-separated' set $S$ of pairs of vertices in $G$, we can always find a path in $G$ that contains every vertex in $S$. For this, we need some definitions of boxes in $G$ and in $G_k$.  We base our definition of `well-separated' on the concept of boxes, which we now define.

@@

We extend these definitions to vertices of a $k\times k$ pseudogrid $G$ with grid-partition $\mathcal{P}:=\{V(P_\mu):\mu\in\VE(G_k)\}$ as follows. For any $\mu\in\VE(G_k)$, define
\[
   \tilde{B}_r(\mu) := \bigcup_{\nu\in \VE(G_r(\mu))} V(P_\nu)
\]
and, for any $v\in V(G)$, let $\tilde{B}_r(v):=\tilde{B}_r(\mu_v)$
where $\mu_v$ denotes the unique element in $\VE(G_k)$ such that $v\in P_{\mu_v}$.\footnote{Technically, the notation for $\tilde{B}_r(v)$ should include the partition $\mathcal{P}$, but we omit this since there will never be any ambiguity as to which partition is being used.}
% For any $v\in V(G)$, define $\tilde{G}_r(v):=G[\tilde{B}_r(v)]$.

\begin{lem}\label{pick_up_two}
  Let $G$ be a $a\times a$ pseudogrid with $a\ge 5$,  let $s,v,w,t$ be vertices of $G$ with $s$ in column $1$ of $G$, $v,w\in\interior_1(G)$, and $t$ in column $a$ of $G$.  Then $G$ contains a path $P$ with endpoints $s$ and $t$ that contains $v$ and $w$.
\end{lem}

The proof of \cref{pick_up_two} is a tedious case analysis that appears in \cref{pick_up_two_proof}.

@@

  Let $X:=X_1\cup X_2\cup X_3$.  By construction, $X$ satisfies (\ref{covers_s}), so it only remains to show that $X$ satisfies (\ref{two_per_box}) and (\ref{disjoint_boxes}).  That $X$ satisfies (\ref{two_per_box}) follows easily from (\ref{disjoint_boxes}), the fact that each element of $S$ is covered by some $x\in X$, and the fact that each $x\in X$ covers at most two elements of $S$.

  To see that $X$ satisifies (\ref{disjoint_boxes}), observe that if $x\in X$ covers $v\in S$ then $B_{p+1}(x)\subseteq B_{2p+2}(v)$. Therefore, if $x\in X$ covers $v\in S$ and $y\in X$ covers $w\in S$ and $B_{p+1}(x)\cap B_{p+1}(y)\neq\emptyset$ then $w\in B_{4p+4}(v)$ or $v\in B_{4p+4}(w)$.  Since $r-1 \ge 4p+4$, this implies that $v=w$ or $v$ and $w$ are an $(r-1)$-pair.  In the former case, $x=y$ and there is nothing to prove. In the latter case, $v,w\in S_2$ or $v,w\in S_3$.  If $v,w\in S_2$ then, again $x=y$ and there is nothing to prove.  If $v,w\in S_3$ then $x$ and $y$ are specifically chosen so that $B_{p+1}(x)$ and $B_{p+1}(y)$ are disjoint.
\end{proof}

@@

\begin{proof}
  Apply \cref{make_disjoint} with $p=\floor{(r-5)/4}$ to obtain the set $X$.  Now, consider a `snake-like' path $P_0$ that contains every row $i$ with $i\equiv 1\pmod{2p+1}$ and, aside from those rows contains only vertices of columns $1$ and $k$.   For each $x\in X$, this path intersects exactly one row of $B_{p+1}(x)$.  For each $x\in X$ we use \cref{pick_up_two} to replace $V(P_0)\cap B_{p+1}(x)$ with a path that is contained in $B_{p+1}(x)$ and contains the (at most two) vertices of $B_{p+1}(x)\cap S$.  After doing this for each $x\in X$ we obtain a path $P$ that contains every vertex in $S$. 
\end{proof}

\todo[inline]{A bunch of $B$'s should be $\tilde{B}$'s}

@@

\begin{proof}
  For simplicity, assume that $G_k$ is embedded in $G_{k+2r}$ with the elements of $Q$ all contained in $\VE(\interior_{2r}(G_{k+2r})$.  In particular, for each $\mu_1\in Q$, this implies that $|B_{\rho}(\nu)| =\vol{\rho}$ for any $\rho\in\{0,\ldots,2r\}$.  (This assumption is not critical, but eliminates the need to discuss boundary conditions that have no effect on the final result.)

@@

\begin{lem}\label{doubled_colour_set}
  For every integer $r\ge 1$ there exists an integer $d\in O(r^4)$ such that the following is true, for every integer $k\ge 1$.
  Let $G$ be a $k\times k$ pseudogrid with grid-partition $\mathcal{P}:=\{V(P_\mu):\mu\in\VE(G_k)\}$ and let $\varphi$ be a vertex colouring of $G$ with the property that, for each $A\subseteq\varphi(V(G))$, $|\varphi_{\mathcal{P}}^{-1}(A)\cap\interior_r(G_k)|\ge d|A|$.
  Then there exists $S\subseteq V(\interior_r(G))$ such that
  \begin{compactenum}[(i)]
    \item \label{hits_both} for each $\alpha\in\varphi(V(G))$, $|\varphi^{-1}(\alpha)\cap S|= 2$; and
    \item \label{spread_out} for each $v\in S$, $|\tilde{B}_r(v)\cap S|\le 2$ for each $v\in S$.
  \end{compactenum}
\end{lem}

\begin{proof}
  We begin by applying \cref{one_colour_per_object} to obtain the colouring $\phi$ of $\VE(G_k)$.  Observe that it is now sufficient to find a $2|\varphi(V(G))|$-element subset $Q\subseteq\VE(\interior_r(G_k))$ such that
  \begin{compactenum}[(i)]
    \item $|\phi^{-1}(\alpha)\cap Q|= 2$ for each $\alpha\in\varphi(V(G))$ and
    \item $|\VE(G_r(\mu))\cap Q|\le 2$ for each $\mu\in Q$.
  \end{compactenum}
  Indeed, with such a $Q$ we can obtain $S$ by taking one vertex of colour $\phi(\mu)$ from $P_\mu$ for each $\mu\in Q$.
